\NAME: A framework for \\ Efficient Intrusion Detection\\ in the Internet of Things

### 2014-10-20 paper.tex
some more rewriting of background

@@

There are different ways to construct a taxonomy for intrusion detection, but
common themes do appear. According to \cite{mishra2004intrusion} and
\cite{ioannis2007towards} there are three major categories: anomaly detection,
signature or misuse detection, and specification-based detection. In
\cite{alrajeh2013intrusion} the authors mostly agree with this topology, but
add the notion of hybrid intrusion detection systems and cross layer intrusion
detection systems as recent advances in research. Although these additional
categories offer very interesting prospects, the authors have to admit that the
impact on the resource-constrained sensor nodes might still be too high.

@@

These taxonomies focus on the act of detecting intrusions themselves. In a
highly distributed environment, an orthogonal categorization, dealing with the
distributed nature of the network, may be even more important.

As in many other situations, a group is often stronger than the sum of its
individual components. Because not all devices are always actively
participating in the network, some might miss clues that would otherwise lead
them to detect intrusions. It is hardly impossible for a single device to
detect a complex attack by itself. Therefore a second important research topic
consists of cooperative algorithms to combine information from single
identities into a group-based decision about alleged
intrusions\cite{zhang2000intrusion}.

A distributed, cooperative algorithm can sometimes be implemented locally. This
is illustrated by the reputation-based detection algorithm introduced in
\cite{ganeriwal2008reputation}, based on the same principle of a watchdog also
found in \cite{mishra2004intrusion} Using this algorithm, nodes can exchange
information about the reputation of other nodes. Combining this information
allows them to decide about their trust in a given node, based on more than
their own, often partial, observations alone.

In \cite{krontiris2009cooperative}, the authors first present a theoretical
foundation to analyze cooperative algorithms. Given these foundations they
continue to present an algorithm consisting of 5 phases. It essentially
implements a voting system based on the Guy Fawkes
protocol\cite{anderson1998new} to identify an intruder. It enables the exchange
of suspected intruder information and allows distributed authentication of
those \emph{votes}. Based on these authenticated votes, a distributed decision
can be made about the commonly identified intruder.

@@

Software attestation is hard and many details can cause an algorithm to fail.
Interesting is the discussion surrounding this topic that was started by
\cite{castelluccia2009difficulty}, in response to the previously mentioned
papers. In \cite{perrig2010refutation} the authors of the original papers
counter many of the objections made to their work, but the general feeling is
that even in the best conditions, there is always a way to circumvent even the
most ingenious algorithm to perform software-based attestation.

@@

Given this classification, we now introduce a design pattern, common to these
algorithms. It will allow us to define a strategy for optimizing the
combination of these algorithms.

From the classification and survey of several algorithms, we can extract two
activities common to most algorithms: the inspection of network traffic and the
evaluation of thresholds.

The former is rather trivial. Network traffic is the carrier along which
malicious payloads reach the device or where patterns of communication can be
observed when other devices in the vicinity are affected by such payloads.

The latter is more complex. An algorithm based on signature or misuse detection
can sometimes undeniably be sure if an attack is real or not. But when dealing
with anomaly detection this line might not be that crisp. Especially in
distributed situations where even a neighbouring device can not be trusted,
detecting intrusion is no absolute science. In this case one has to rely on
statistics and thresholds to decide when to drop trust in a given device.

Simply checking a threshold after updating a counter, is also not always a good
strategy. A simple example is that of a watchdog\cite{mishra2004intrusion},
observing the actions of other devices. There are valid reasons why a devices
might miss its theoretical window in which certain actions are expected. Simply
reacting on such basic threshold might result in false positives. A way to
overcome this is to decouple the detection of problems and the decision what to
do about it.

@@

We now extend the scope of the design pattern, to include its application. When
implementing different algorithms, one will typically encounter a pattern such
as describes in listing \ref{alg:id-algo-application}.

@@

The different implementations would be linked and called in a sequential way,
when new messages are observed on the network and at a regular interval to
allow them to do internal housekeeping.

@@

Reimplementing the algorithms to eliminate these issues is not a valid option.
This development cost typically isn't a one-time cost, e.g. in case a new
version of the algorithm is released, or simply when a different set of
algorithms is selected due to a changing risk analysis. Finally, changing, or
simply implementing the algorithms from scratch, also holds the risk of making
mistakes.

\subsection{Network Usage}

@@

The heterogeneity of the IoT devices, results in many different software
stacks, each with their own network API. This also explains why there are very
few implementations readily available to developers, their is no common
platform to target.

@@

Listing \ref{lst:heartbeat} illustrates some of the features mentioned above,
using a small but representative example of an algorithm: a heartbeat. Using a
heartbeat, nodes can broadcast a message at a regular interval, allowing other
nodes to monitor its availability. Missing heartbeats could indicate problems,
such as physical capturing of a certain node.

\lstinputlisting[
  float,
  basicstyle=\small\ttfamily,
  label=lst:heartbeat,
  caption=Heartbeat in \NAME
]{src/heartbeat.foo}

>>>>

These two topics lead to the first two classes: detecting intrusions and
cooperative decision making. We also identify software attestation as a third
class. Although it can be considered as a special case of a combination of both
other classes, its focus is not on the detection of intrusions, but on the
effect of an intrusion on the system itself.

@@

There are different ways to construct a taxonomy for intrusion detection, but
common themes do appear. According to \cite{mishra2004intrusion} and
\cite{ioannis2007towards} there are three major categories: anomaly detection,
signature or misuse detection, and specification-based detection. In
\cite{alrajeh2013intrusion} the authors mostly agree with this topology, but
add the notion of hybrid intrusion detection systems and cross layer intrusion
detection systems as recent advances in research. Although these additional
categories offer very interesting prospects, the authors have to admit that the
impact on the resource-constrained devices might still be too much. We
therefore focus on the three categories agreed upon by most sources.

\begin{LaTeXdescription}

  \item[Anomaly detection] considers a baseline profile and evaluates all
  activity according to this profile. Aberrant behavior is flagged as a
  possible intrusion, but this is no exact science. The baseline profile can
  evolve over time, especially in wireless and mobile environments. The flagged
  incidents should therefore also be handled with care and no absolute
  decisions can be made from single events.

  \item[Signature or misuse detection] starts with known patterns of attacks
  and tries to find evidence in the actions of and communication with a device.
  Because many attacks use valid ways to communicate with services on the
  devices, this can also produce false positives and offers no black and white
  certainty.

  \item[Specification-based detection] uses constraints that describe the
  correct behavior of the system. Although that this approach can detect
  previously unknown attacks, it can sometimes not detect specific attacks that
  can be detected by anomaly or misuse detection.

\end{LaTeXdescription}

@@

These taxonomies focus on the act of detecting intrusions themselves. In a
highly distributed environment, an orthogonal categorization, dealing with the
distributed nature of the network and its ways of communicating, may be even
more important.

As in many other situations, a group is often stronger than the sum of its
individual components. Because not all devices are always actively
participating in the network, some might miss clues that would otherwise lead
them to detect intrusions. It is hardly impossible for a single device to
detect a complex attack by itself. Therefore a second important research topic
consists of cooperative algorithms to combine information from single
identities into a group-based decision about alleged
intrusions\cite{zhang2000intrusion}. We identify two ways by which this
cooperative decision making can be constructed and focus on the way
participants exchange information: broadcasting and interactive.

\begin{LaTeXdescription}

  \item[Broadcasting] can sometimes allow a distributed, cooperative algorithm
  to be implemented locally. This is illustrated by the reputation-based
  detection algorithm introduced in \cite{ganeriwal2008reputation}, based on
  the same principle of a watchdog also found in \cite{mishra2004intrusion}
  Using this algorithm, nodes can exchange information about the reputation of
  other nodes, using only broadcasted messages. Combining this information
  allows them to decide about their trust in a given node, based on more than
  their own, often partial, observations alone.

  \item[Interactive] communication allows algorithms to exchange information to
  reach a consensus. In \cite{krontiris2009cooperative}, the authors first
  present a theoretical foundation to analyze cooperative algorithms. Given
  these foundations they continue to present an algorithm consisting of 5
  phases. It essentially implements a voting system based on the Guy Fawkes
  protocol\cite{anderson1998new} to identify an intruder. It enables the
  exchange of suspected intruder information and allows distributed
  authentication of those \emph{votes}. Based on these authenticated votes, a
  distributed decision can be made about the commonly identified intruder.

\end{LaTeXdescription}

@@

Software attestation is hard and many details can cause an algorithm to fail. A
discussion surrounding this topic, started by \cite{castelluccia2009difficulty}
in response to the previously mentioned papers, illustrates this. In
\cite{perrig2010refutation} the authors of the original papers counter many of
the objections made to their work, but the general feeling is that even in the
best conditions, there is always a way to circumvent the most ingenious
algorithm to perform software-based attestation.

Within the context of this paper, we will not consider software attestation as
a special class and exclude it from our scope, only focusing on detecting
intrusions and distributed decision making.

Figure \ref{fig:classification} gives a visual overview of the classification
presented above.

\begin{figure}[!ht]
  \centering
  \scalebox{.85}{
\begin{tikzpicture}[
  sibling distance=20mm,node distance=40mm
  ]
  \node (detection)             [align=center] {Detecting\\Intrusions}
    child {node (anomaly)       [align=center] {Anomaly}}
    child {node (signature)     [align=center] {Signature\\Misuse}}
    child {node (specification) [align=center] {Specification}};
  \node (decision)              [align=center,right=of detection] {Decision\\Making}
    child {node (broadcasting)  [align=center] {Broadcasting}}
    child {node (interactive)   [align=center] {Interactive}};
\end{tikzpicture}
}
  \caption{Classification of ID Algorithms}
  \label{fig:classification}
\end{figure}

@@

Given this classification, we now introduce a design pattern that identifies
common aspects of these algorithms. It will allow us to define a strategy for
optimizing the combination of these algorithms. From the classification and
survey of several algorithms, we extract two activities common to most
algorithms: the inspection of network traffic and the evaluation of thresholds.

The former is trivial: network traffic is the carrier along which malicious
payloads reach the device or where patterns of communication can be observed
when other devices in the vicinity are affected by such payloads.

The latter is more complex: an algorithm based on signature or misuse detection
maybe sometimes can undeniably decide if an attack is real or not. But when
dealing with anomaly detection this line surely isn't that crisp. Especially in
distributed situations where even a neighboring device can not be trusted,
detecting intrusion is no absolute science. In this case one has to rely on
statistics and thresholds to decide when to drop trust in a given device.

Simply checking a threshold after updating some counter, is also not always a
good strategy. A simple example is that of a
watchdog\cite{mishra2004intrusion}, observing the actions of other devices.
There are valid reasons why a devices might miss the theoretical window in
which certain actions are expected. Simply reacting on such basic threshold
might result in false positives. A way to overcome this is to decouple the
detection of problems and the decision what to do about them.

@@

We now extend the scope of the design pattern, to include its application in
the context of multiple algorithms. When implementing different algorithms, one
will typically encounter a pattern such as describes in listing
\ref{alg:id-algo-application}.

@@

The different implementations would be linked and called in a sequential way,
when new messages are observed on the network and at a regular interval to
allow them to do internal housekeeping, such as threshold evaluation.

@@

Reimplementing the algorithms manually to eliminate these issues is not a valid
option. This development cost typically isn't a one-time cost, e.g. in case a
new version of the algorithm is released, or simply when a different set of
algorithms is selected due to a changing risk analysis. Finally, changing, or
simply implementing the algorithms from scratch, also holds the risk of making
mistakes.

\subsection{Network Usage and Heterogeneity}

@@

The heterogeneity of the IoT devices, results in many different software
stacks, each with their own network API. This also explains why there are very
few implementations readily available to developers: there is no common
platform to target for researchers.

@@

Listing \ref{lst:watchdog} illustrates some of the features mentioned above,
using a small but representative example of an algorithm: a
watchdog\cite{mishra2004intrusion}. Using a heartbeat, nodes can broadcast a
message at a regular interval, allowing other nodes to monitor its
availability. Missing heartbeats could indicate problems, such as physical
capturing of a certain node.

\lstinputlisting[
  float,
  basicstyle=\small\ttfamily,
  label=lst:watchdog,
  caption=Heartbeat in \NAME
]{src/watchdog.foo}
